---
title: "W241 Final Project | Decoding Game of Thrones"
subtitle: "The Effect of Gender Identity on Learning Outcomes"
author: "Alex West, Anna Jacobson, and Apik Zorian"
date: "12/10/19"
output: 
    pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_format = c('pdf_document')) 
    })
---


```{r message=FALSE, include=FALSE, results='hide'}
# load packages 
library(data.table)
library(foreign)
library(lmtest)
library(sandwich)
library(data.table)
library(stargazer)
library(ggplot2)
library(dplyr)
library(Hmisc)
library(knitr)
library(pwr)
library(stringr)
```

```{r Functions, include=FALSE}

# calculate effect size
effect <- function(g1,g2) {
  effect_size <- ((mean(g1)-mean(g2))/(sd(c(g1, g2))))
  return(effect_size)
}

# calculate robust standard errors
rse <- function(m) {
  m$vcovHC_ <- vcovHC(m)
  robust_standard_error <- sqrt(diag(m$vcovHC_))
  return(robust_standard_error)
}

# output two-model regression table using robust standard errors
reg_rse_2 <- function (m1, m2, name) {
    stargazer(m1, m2,
          type = "latex",
          se = list(rse(m1),rse(m2)),
          add.lines = list(
            c('Type of Std. Error', 'Robust', 'Robust')
          ),
          header = F,
          title = name,
          digits = 3)
}

# output three-model regression table using robust standard errors
reg_rse_3 <- function (m1, m2, m3, name) {
    stargazer(m1, m2, m3,
          type = "latex",
          se = list(rse(m1),rse(m2),rse(m3)),
          add.lines = list(
            c('Type of Std. Error', 'Robust', 'Robust', 'Robust')
          ),
          header = F,
          title = name,
          digits = 3)
}

# output four-model regression table using robust standard errors
reg_rse_4 <- function (m1, m2, m3, m4, name) {
    stargazer(m1, m2, m3, m4,
          type = "latex",
          se = list(rse(m1),rse(m2),rse(m3),rse(m4)),
          add.lines = list(
            c('Type of Std. Error', 'Robust', 'Robust', 'Robust', 'Robust')
          ),
          header = F,
          title = name,
          digits = 3)
}

# output five-model regression table using robust standard errors
reg_rse_5 <- function (m1, m2, m3, m4, m5, name) {
    stargazer(m1, m2, m3, m4, m5,
          type = "latex",
          se = list(rse(m1),rse(m2),rse(m3),rse(m4),rse(m5)),
          add.lines = list(
            c('Type of Std. Error', 'Robust', 'Robust', 'Robust', 'Robust', 'Robust')
          ),
          header = F,
          title = name,
          omit.stat=c("f", "ser"),
          digits = 3,
          font.size = "small",
          column.sep.width = "-5pt")
}

```

```{r Initial Data Loading, include=FALSE}
e1 <- fread("./Data_Clean_PhI_191123_09.48.csv")
e2 <- fread("./Data_Clean_PhII_191125_18.16.csv")
dim(e1)
dim(e2)
``` 

##3.1.4 | Experiment Design 1: Two-Group Model  

###Individual Values Plot

```{r fig.height=4}
ED1_IVP <- ggplot(e1, aes(x = as.factor(Same_Gen), y = Total_Scaled)) +
  geom_boxplot(color="black", fill="gray", position=position_dodge(1)) + 
  stat_boxplot(geom ='errorbar') +
  geom_jitter(aes(x = as.factor(Same_Gen), y = Total_Scaled, fill=as.factor(G_s)), 
                size=3, color="black", pch=21, position=position_jitter(0.1)) +
  scale_x_discrete(labels= c("Same-Gender Instructor", "Opposite-Gender Instructor")) +
  labs(title = "Experiment Design 1: Two-Group", subtitle = "All Subjects") +
  labs(y = "Total Phase I Score (Scaled)", x = "Treatment Condition") + 
  scale_fill_discrete(name = "Student Gender", labels = c("Female", "Male"))

ggsave(filename="ED1_IVP.png", plot=ED1_IVP, width = 7, height = 4, units = "in")
```

### Hypothesis Testing

```{r}
#calculate group distributions
same_gender_group <- e1[Same_Gen == 0, Total_Scaled, ]
mixed_gender_group <- e1[Same_Gen == 1, Total_Scaled, ]
```

```{r}
t.test(same_gender_group, mixed_gender_group)
pwr.t2n.test(n1=length(same_gender_group), n2=length(mixed_gender_group), d=effect(same_gender_group,mixed_gender_group))
```

##3.1.5 | Experiment Design 2: Four-Group Model

###Individual Values Plot

```{r}
ED2_IVP <- ggplot(e1, aes(x = as.factor(PA_2), y = Total_Scaled)) + 
  geom_boxplot(color="black", fill="gray", position=position_dodge(1)) + 
  stat_boxplot(geom ='errorbar') +
  geom_jitter(aes(x = as.factor(PA_2), y = Total_Scaled, fill=as.factor(G_s)), 
              size=3, color="black", pch=21, position=position_jitter(0.1)) +
  scale_x_discrete(labels= c("FF", "FM", "MF", "MM")) +
  labs(title = "Experiment Design 2: Four-Group", subtitle = "All Subjects") + 
  labs (y = "Total Phase I Score (Scaled)", 
        x = "Treatment Condition") + 
  scale_fill_discrete(name = "Student Gender", labels = c("Female", "Male"))

ggsave(filename="ED2_IVP.png", plot=ED2_IVP, width = 7, height = 4, units = "in")
```

### Hypothesis Testing

```{r}
#calculate group distributions
FF_group <- e1[FF == 1, Total_Scaled, ]
FM_group <- e1[FM == 1, Total_Scaled, ]
MF_group <- e1[MF == 1, Total_Scaled, ]
MM_group <- e1[MM == 1, Total_Scaled, ]
```

```{r}
t.test(FF_group, FM_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(FF_group), d=effect(FF_group, FM_group), type = "paired")
```

```{r}
t.test(MF_group, MM_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(MF_group), d=effect(MF_group, MM_group), type = "paired")
```


##3.1.6 | Experiment Design 3: Eight-Group Model

###Individual Values Plot

```{r fig.height=4}
ED3_IVP <- ggplot(e1, aes(x = as.factor(PA_3), y = Total_Scaled)) + 
  geom_boxplot(color="black", fill="gray", position=position_dodge(1)) + 
  stat_boxplot(geom ='errorbar') +
  geom_jitter(aes(x = as.factor(PA_3), y = Total_Scaled, 
                  fill=as.factor(G_s)), size=3, color="black", pch=21, 
              position=position_jitter(0.1)) +
  scale_x_discrete(labels= c("FFA", "FMB","FMA", "FFB", "MFA", "MMB","MMA", "MFB")) +
  labs(title = "Experiment Design 3: Eight-Group", subtitle = "All Subjects") + 
  labs (y = "Total Phase I Score (Scaled)", 
        x = "Treatment Condition") + 
  scale_fill_discrete(name = "Student Gender", labels = c("Female", "Male"))

ggsave(filename="ED3_IVP.png", plot=ED3_IVP, width = 7, height = 4, units = "in")
```

### Hypothesis Testing

```{r}
#calculate group distributions
FFA_group <- e1[FFA == 1, Total_Scaled, ]
FMB_group <- e1[FMB == 1, Total_Scaled, ]
FMA_group <- e1[FMA == 1, Total_Scaled, ]
FFB_group <- e1[FFB == 1, Total_Scaled, ]
MFA_group <- e1[MFA == 1, Total_Scaled, ]
MMB_group <- e1[MMB == 1, Total_Scaled, ]
MMA_group <- e1[MMA == 1, Total_Scaled, ]
MFB_group <- e1[MFB == 1, Total_Scaled, ]
Not_FFA_group <- e1[FFA != 1, Total_Scaled, ]
```

```{r}
t.test(FFA_group, FMB_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(FFA_group), d=effect(FFA_group,FMB_group), type = "paired")
```

```{r}
t.test(FMA_group, FFB_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(FMA_group), d=effect(FMA_group,FFB_group), type = "paired")
```

```{r}
t.test(MFA_group, MMB_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(MFA_group), d=effect(MFA_group,MMB_group), type = "paired")
```

```{r}
t.test(MMA_group, MFB_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(MMA_group), d=effect(MMA_group,MFB_group), type = "paired")
```


```{r}
t.test(FFA_group, Not_FFA_group)
pwr.t2n.test(n1=length(FFA_group), n2=length(Not_FFA_group), d=effect(FFA_group,Not_FFA_group))
```

### Regression Analysis

```{r, results='asis'}
# fit regression models
m3c1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e1)
m3c2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e1)

# output results
reg_rse_2(m3c1, m3c2, "Experiment Design 3: Eight-Group Models")
```

```{r, results='asis'}
# fit regression models
m3f1 <- lm(Total_Scaled ~ FFA, data = e1)
m3f2 <- lm(Total_Scaled ~ FFA + IFE_Scaled, data = e1)

# output results
reg_rse_2(m3f1, m3f2, "Experiment Design 3: Female Students/Female Instructor/Video A")
```

##4.1 | Secondary Analyses 

### Student Subgroup

```{r}
# create student subset 
e_student <- subset(e1, e == 1)
dim(e_student)
```

```{r, results='asis'}
# fit regression models
m4a1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_student)
m4a2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e_student)
m4a3 <- lm(Total_Perf ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_student)
```

### Under 30 Subgroup

```{r}
# create under 30 subset 
e_young <- subset(e1, b == 1 | b == 2)
dim(e_young)
```

```{r, results='asis'}
# fit regression models
m4b1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_young)
m4b2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e_young)
m4b3 <- lm(Total_Perf ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_young)
```

### Over 30 Subgroup

```{r}
# create over 30 subset 
e_old <- subset(e1, b == 3 | b == 4 | b == 5)
dim(e_old)
```

```{r, results='asis'}
# fit regression models
m4c1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_old)
m4c2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e_old)
m4c3 <- lm(Total_Perf ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_old)
```

### Highly Educated Subgroup

```{r}
# create highly educated subset 
e_edu <- subset(e1, f == 6 | f == 7)
dim(e_edu)
```

```{r, results='asis'}
# fit regression models
m4d1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_edu)
m4d2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e_edu)
m4d3 <- lm(Total_Perf ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_edu)
```

### Native English Speaker Subgroup

```{r}
# create native English speaker subset 
e_eng <- subset(e1, d == 1)
dim(e_eng)
```

```{r, results='asis'}
# fit regression models
m4e1 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_eng)
m4e2 <- lm(Total_Scaled ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA + IFE_Scaled, data = e_eng)
m4e3 <- lm(Total_Perf ~ FFA + FMB + FMA + FFB + MFA + MMB + MMA, data = e_eng)
```

### Subgroup Results

```{r, results='asis'}
# output results
reg_rse_5(m4a2, m4b2, m4c2, m4d2, m4e2, "Subgroup Analysis")
```

##Appendix B

###B.1 | Exploratory Data Analysis: Scores


```{r}
EDA_Total_Scaled <- ggplot(e1, aes(x=as.factor(Total_Scaled), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  labs(title = "Total Score", subtitle = "Phase I Subjects") + 
  labs(x = "Total Score (Scaled to 100)", y = "Count") +
  guides(fill=FALSE)
  
EDA_Total_Scaled

ggsave(filename="EDA_Total_Scaled.png", plot=EDA_Total_Scaled, width = 7, height = 4, units = "in")
```

```{r}
EDA_Recall_Scaled <- ggplot(e1, aes(x=as.factor(Recall_Scaled), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  labs(title = "Recall Score", subtitle = "Phase I Subjects") + 
  labs(x = "Recall Score (Scaled to 100)", y = "Count") +
  guides(fill=FALSE)
  
EDA_Recall_Scaled

ggsave(filename="EDA_Recall_Scaled.png", plot=EDA_Recall_Scaled, width = 7, height = 4, units = "in")
```

```{r}
EDA_Comp_Scaled <- ggplot(e1, aes(x=as.factor(Comp_Scaled), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  labs(title = "Comprehension Score", subtitle = "Phase I Subjects") + 
  labs(x = "Comprehension Score (Scaled to 100)", y = "Count") +
  guides(fill=FALSE)
  
EDA_Comp_Scaled

ggsave(filename="EDA_Comp_Scaled.png", plot=EDA_Comp_Scaled, width = 7, height = 4, units = "in")
```

```{r}
EDA_Ret_Scaled <- ggplot(e2, aes(x=as.factor(Ret_Scaled), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  labs(title = "Retention Score", subtitle = "Phase II Subjects") + 
  labs(x = "Retention Score (Scaled to 100)", y = "Count") +
  guides(fill=FALSE)
  
EDA_Ret_Scaled

ggsave(filename="EDA_Ret_Scaled.png", plot=EDA_Ret_Scaled, width = 7, height = 4, units = "in")
```

###B.2 | Exploratory Data Analysis: Covariates

```{r}
EDA_b <- ggplot(e1, aes(x=as.factor(b), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  scale_x_discrete(labels = c("Under 21", "21-30", "31-40", "41-50", "Over 50")) +
  labs(title = "Covariate: Subject Age", subtitle = "Phase I Subjects") + 
  labs(x = "Age Group (Years)", y = "Count") +
  guides(fill=FALSE)
  
EDA_b

ggsave(filename="EDA_b.png", plot=EDA_b, width = 7, height = 4, units = "in")
```

```{r}
EDA_c <- ggplot(e1, aes(x=as.factor(c), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  scale_x_discrete(labels = c("White", "Black", "Asian", "Other", "Biracial")) +
  labs(title = "Covariate: Subject Ethnicity", subtitle = "Phase I Subjects") + 
  labs(x = "Ethnicity", y = "Count") +
  guides(fill=FALSE)
  
EDA_c

ggsave(filename="EDA_c.png", plot=EDA_c, width = 7, height = 4, units = "in")
```

```{r}
EDA_d <- ggplot(e1, aes(x=as.factor(d), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  scale_x_discrete(labels = c("First Language", "Second Language", "Third or More Language")) +
  labs(title = "Covariate: Subject English Acquisition", subtitle = "Phase I Subjects") + 
  labs(x = "English Acquisition", y = "Count") +
  guides(fill=FALSE)
  
EDA_d

ggsave(filename="EDA_d.png", plot=EDA_d, width = 7, height = 4, units = "in")
```

```{r}
EDA_e <- ggplot(e1, aes(x=as.factor(e), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  scale_x_discrete(labels = c("Yes", "No")) +
  labs(title = "Covariate: Subject Student Status", subtitle = "Phase I Subjects") + 
  labs(x = "Current Student", y = "Count") +
  guides(fill=FALSE)
  
EDA_e

ggsave(filename="EDA_e.png", plot=EDA_e, width = 7, height = 4, units = "in")
```

```{r}
EDA_f <- ggplot(e1, aes(x=as.factor(f), fill = ..count..)) + 
  geom_bar(color = "black", position = "identity") +
  scale_fill_gradient("Count",low = "white", high = "gray") +
  scale_x_discrete(labels = str_wrap(c("Less than high school", 
                                       "High school graduate",
                              "Some college", "4 year degree", 
                              "Master or Professional degree", 
                              "Doctorate"), width=10)) +
  labs(title = "Covariate: Subject Education", subtitle = "Phase I Subjects") + 
  labs(x = "Level of Education Completed", y = "Count") +
  guides(fill=FALSE)
  
EDA_f

ggsave(filename="EDA_f.png", plot=EDA_f, width = 7, height = 4, units = "in")
```

###B.3 | Experiment Design 1: Two-Group Model

```{r, results='asis'}
# fit regression models
m2a1 <- lm(Total_Scaled ~ Same_Gen, data = e1)
m2a2 <- lm(Total_Scaled ~ Same_Gen + IFE_Scaled, data = e1)

# output results
reg_rse_2(m2a1, m2a2, "Experiment Design 1: Two Groups")
```


###B.4 | Experiment Design 2: Four-Group Model

```{r, results='asis'}
# fit regression models
m2c1 <- lm(Total_Scaled ~ FF + FM + MF, data = e1)
m2c2 <- lm(Total_Scaled ~ FF + FM + MF + IFE_Scaled, data = e1)

# output results
reg_rse_2(m2c1, m2c2, "Experiment Design 2: Four Groups")
```

###3.2 | Subgroup Analyses

### Student Subgroup

```{r, results='asis'}
# output results
reg_rse_3(m4a1, m4a2, m4a3, "Subgroup Analysis: Students")
```

### Under 30 Subgroup

```{r, results='asis'}
# output results
reg_rse_3(m4b1, m4b2, m4b3, "Subgroup Analysis: Under 30 Years Old")
```

### Over 30 Subgroup

```{r, results='asis'}
# output results
reg_rse_3(m4c1, m4c2, m4c3, "Subgroup Analysis: Over 30 Years Old")
```

### Highly Educated Subgroup

```{r, results='asis'}
# output results
reg_rse_3(m4d1, m4d2, m4d3, "Subgroup Analysis: Highly Educated")
```


### Native English Speaker Subgroup

```{r, results='asis'}
# output results
reg_rse_3(m4e1, m4e2, m4e3, "Subgroup Analysis: Native English Speakers")
```

###Analysis: Video A vs. B

```{r}
#calculate group distributions
VideoA_group <- e1[Video == 1, Total_Scaled, ]
VideoB_group <- e1[Video == 2, Total_Scaled, ]
```

```{r}
t.test(VideoA_group, VideoB_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(VideoA_group), d=effect(VideoA_group, VideoB_group), type = "paired")
```

###Analysis: Recall, Comprehension, and Retention

```{r}
#calculate group distributions
Total_group <- e1[ , Total_Scaled, ]
Recall_group <- e1[, Recall_Scaled, ]
Comp_group <- e1[, Comp_Scaled, ]
Ret_group <- e2[ , Ret_Scaled, ]
```

```{r}
t.test(Recall_group, Total_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(Recall_group), d=effect(Recall_group, Total_group), type = "paired")
```

```{r}
t.test(Comp_group, Total_group, paired = TRUE, alternative = "two.sided")
pwr.t.test(n=length(Comp_group), d=effect(Comp_group, Total_group), type = "paired")
```

```{r}
t.test(Ret_group, Total_group)
pwr.t2n.test(n1=length(Ret_group), n2=length(Total_group), d=effect(Ret_group, Total_group))
```
